CSCI 580 : Monte Carlo Path Tracing

### 2015-12-03 paper/paper.tex
reorder paragraph + beginning of slides
finish powerpoint and talk about multithread in paper

\begin{description}
  \item[1)] Generate a ray from the camera (position x,y, horizontal and vertical fields of view u,v), passing through the image plane, with $weight = 1$
  \item[2)] Find the point of intersection with the nearest surface
  \item[3)] Determine whether the surface is a light source or not (as indicated by a special coefficients for light emittance)
  \item[3A)] if it is a light source, return $weight * light$
  \item[3B)] if not, $weight \mathrel{*}= \frac{Ks+Kd+Ka}{3}$ and randomly scatter a new ray according to a probability density function (pdf) approximating its bidirectional refletance distribution function (brdf), and go to step 2
\end{description}

@@

\section{Graphics Pipeline}
The graphics pipeline was designed as follows: initialization is carried out, wherein global properties such as camera position and lights are defined. Then, the meshes are loaded into memory as a data structure encompassing all triangles. Each vertex in the triangle has a position and a normal in world space coordinates, as well as information on their specular, diffuse and ambient coefficients, which differs from the global Ks, Kd and Ka found in the assignment renderer. Also associated with vertices is a lighting coefficient for surfaces which emit light. Next, this set of all geometry is used by the path tracing technique to determine colour at each pixel, which is finally fed to the same display as the one used in previous asignments.

@@

@@

\par
The image becomes sharp at around $s=10$ for any value of $d$, and artifacts persist up to $s=40, d=4$ and $s=100,d=1$.

@@

\centering
\includegraphics[width=2.5in]{10s_1d_5m42s}
\caption{10 sets of rays, recursion depth 1, time: 5m42s}
\label{10s_1d_5m42s}

@@

\centering
\includegraphics[width=2.6in]{20s_1d_11m26s}
\caption{20 sets of rays, recursion depth 1, time: 11m26s}
\label{20s_1d_11m26s}

@@

\centering
\includegraphics[width=2.5in]{100s_1d_40m21s}
\caption{100 sets of rays, recursion depth 1, time: 40m21s}
\label{100s_1d_40m21s}

@@

\centering
\includegraphics[width=2.6in]{100s_16d_42m10s}
\caption{100 sets of rays, recursion depth 1, time: 42m10s}
\label{1100s_16d_42m10s}

@@

\section{Challenges \& Observations}
There were many obstacles on the way to the completion of the project. Firstly, the codebase of the assignments is not particularly suited for ray casting: rather, it is suited for processing the geometry one triangle at a time since it rasterized them. Thus, a new architectured had to be written to replace the old one, which was more of a time sink than first anticipated. 
\par
Second, building the scene was a long and tedious process, as the geometry had to be written in the ASCII format as read by the assignment renderer. Exacerbating this process is that path tracing is an expensive algorithm: to test the placement of geometry, the addition of features in the application or to debug it, one must wait between half a minute up to an hour depending on the quality of the generated image.
\par
Additionally, until a late phase of the project, we opted to continue to use the teapot that was featured in all previous assignments. However, since the teapot contains a great number of triangles, it was  expensive to calculate which triangle intersecting with the ray was the closest, since it required iterating through all the triangles. Since this operation was conducted per ray cast, the run-time cost of rendering the teapot was very high.

@@

\par
Another opportunity for extension of the application is in its performance. There are two main avanues for this: improvement of runtime complexity and parallelization. In terms of runtime complexity, for determining the closest triangle intersected, one could make use of binary space partition trees rather than a brute force approach. Also, bounding volumes could be used to test for intersection with objects with complex geometries, such as the teapot, which would be encapsulated in a much more simple box shape for collision testing.

\par
As for parallelization, the ray casting approach lends itself well to being parallelized since each ray is reflected irrespective of rays casting through the same pixels (the samples), as well as irrespectively of the rays being cast through other pixels. Therefore, each individual ray could be calculated simultaneously. Incidentally, this process lends itself well to GPUs, which specialize in computing large sets of similar data.

@@

\par
Furthermore, for images where $s \ge 10$, fully dark pixels are displayed in a noticeable, somewhat circular pattern. This could be the result of the uniform random reflectance along the semi-circular hemisphere, although there is currently not a definitive answer to why these artifacts appear.

>>>>

\begin{description}
  \item[1)] Generate a ray from the camera (position x,y, horizontal and vertical fields of view u,v), passing through the image plane, with $weight = 1$
  \item[2)] Find the point of intersection with the nearest surface
  \item[3)] Determine whether the surface is a light source or not (as indicated by a special coefficients for light emittance)
  \item[3A)] if it is a light source, return $weight * light$
  \item[3B)] if not, $weight \mathrel{*}= \frac{Ks+Kd+Ka}{3}$ and randomly scatter a new ray according to a probability density function (pdf) approximating its bidirectional reflectance  distribution function (brdf), and go to step 2
\end{description}

@@

\section{Graphics Pipeline}
The graphics pipeline was designed as follows: initialization is carried out, wherein global properties such as camera position and lights are defined. Then, the meshes are loaded into memory as a data structure encompassing all triangles. Each vertex in the triangle has a position and a normal in world space coordinates, as well as information on their specular, diffuse and ambient coefficients, which differs from the global Ks, Kd and Ka found in the assignment renderer. Also associated with vertices is a lighting coefficient for surfaces which emit light. Next, this set of all geometry is used by the path tracing technique to determine colour at each pixel, which is finally fed to the same display as the one used in previous assignments.

@@

\par
As for parallelization, the ray casting approach lends itself well to being parallelized since each ray is reflected irrespective of rays casting through the same pixels (the samples), as well as irrespectively of the rays being cast through other pixels. Therefore, multithreaded was implemented, where individual ray could be calculated simultaneously

@@

\par
Multithreading was mostly not used in the generation of the images, so that times could better be compared, although for high values of $s$ images were generated with multithreading. Since the application uses 25 threads, we see a decrease in speed by 25 times!

\par
The image becomes sharp at around $s=10$ for any value of $d$, but artifacts persist all the way up to $s=200$, $d=16$.

@@

\centering
\includegraphics[width=2.6in]{10s_1d_5m42s}
\caption{10 sets of rays, recursion depth 1, time: 5m42s}
\label{10s_1d_5m42s}

@@

\centering
\includegraphics[width=2.5in]{20s_1d_11m26s}
\caption{20 sets of rays, recursion depth 1, time: 11m26s}
\label{20s_1d_11m26s}

@@

\centering
\includegraphics[width=2.6in]{100s_1d_40m21s}
\caption{100 sets of rays, recursion depth 1, time: 40m21s. Time with multithreading: 2min52s}
\label{100s_1d_40m21s}

@@

\centering
\includegraphics[width=2.5in]{100s_16d_42m10s}
\caption{100 sets of rays, recursion depth 16, time: 42m10s. Time with multithreading: 2min59s}
\label{100s_16d_42m10s}

@@

\centering
\includegraphics[width=2.5in]{200s_16d_5min53s}
\caption{200 sets of rays, recursion depth 16, time with multithreading: 5min53s}
\label{200s_16d_5min53s}

\section{Challenges \& Observations}
There were many obstacles on the way to the completion of the project. Firstly, the codebase of the assignments is not particularly suited for ray casting: rather, it is suited for processing the geometry one triangle at a time since it rasterized them. Thus, a new architecture had to be written to replace the old one, which was more of a time sink than first anticipated. 
\par
Second, building the scene was a long and tedious process, as the geometry had to be written in the ASCII format as read by the assignment renderer. Exacerbating this process is that path tracing is an expensive algorithm: to test the placement of geometry, the addition of features in the application or to debug it, one must wait between half a minute up to an hour depending on the quality of the generated image.
\par
Additionally, until a late phase of the project, we opted to continue to use the teapot that was featured in all previous assignments. However, since the teapot contains a great number of triangles, it was  expensive to calculate which triangle intersecting with the ray was the closest, since it required iterating through all the triangles. Since this operation was conducted per ray cast, the run-time cost of rendering the teapot was very high.
\par
Furthermore, for images where $s \ge 10$, fully dark pixels are displayed in a noticeable, somewhat circular pattern. This could be the result of the uniform random reflectance along the semi-circular hemisphere, although there is currently not a definitive answer to why these artifacts appear.

@@

\par
Another opportunity for extension of the application is in its performance. In terms of runtime complexity, for determining the closest triangle intersected, one could make use of binary space partition trees rather than a brute force approach. Also, bounding volumes could be used to test for intersection with objects with complex geometries, such as the teapot, which would be encapsulated in a much simpler box shape for collision testing.

\par
Although the application does use multithreading, the algorithm could also lend itself well to a GPU, which specializes in computing large sets of similar data.

@@
